\Title

### 2013-06-06 synmut.tex
a few more little changes

\begin{abstract}
\noindent
Intrapatient HIV-1 evolution is driven by the adaptive immune system
resulting in rapid change of HIV-1 proteins. When cytotoxic CD8${}^+$ T-cells
or neutralizing antibodies target a new epitope, the virus often escapes via
nonsynonymous mutations that impair recognition. Synonymous mutations do not
affect this interplay and are often assumed to be neutral. We test this
assumption by tracking the abundance of synonymous mutations in 
longitudinal intrapatient data from the \shankaregion{} part of the \env{}
gene. We find that most synonymous variants are lost even though
they often reach high frequencies in the viral population, suggesting a
cost to the virus. Using published data
from SHAPE assays, we find that synonymous mutations that disrupt base pairs
in RNA stems flanking the variable loops of gp120 are more likely to be lost than other
synonymous changes; these RNA hairpins might be important for HIV-1.
Computational modeling indicates that (i) most synonymous mutations in this
genomic region have a selection coefficient of the order of $-0.002$, and
(ii) synonymous variants rise to high frequency by genetic hitchhiking on neighboring beneficial
variants. This weak selection against synonymous substitutions does not
result in a strong pattern of conservation in cross sectional data, but
slows down the rate of evolution considerably. Our findings are consistent with the
notion that large scale patterns of RNA structure are functionally
relevant, while the precise base pairing pattern is not.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
HIV-1 evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via cytotoxic CD8${}^+$ T-cells (CTLs) and neutralizing antibodies
(nAbs)~\citep{rambaut_causes_2004} and facilitated by HIV-1's high mutation rate
~\citep{mansky_lower_1995,abram_nature_2010}. Escape mutations
in epitopes targeted by CTLs are typically observed during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving parts of the HIV-1 genome are the variable
loops V1-V5 in the envelope protein gp120 (V loops), which change to avoid recognition by
nAbs. Escape mutations in \env{}, the gene encoding gp120, spread through the
viral population within a few months.
Consistent with this time scale, it is found that serum from a particular time
typically neutralizes autologous virus extracted more than 3-6 months earlier but not contemporary
virus \citep{richman_rapid_2003}.

@@

Most positions are only transiently variable and variants will either fix or
will be lost -- at least in small samples. Given an SNV is at a certain
frequency $\nu$, the probability of fixation is higher for beneficial
SNVs than for neutral ones; in turn, neutral variants fix more frequently than
deleterious ones. The fixation probability of a
neutral SNV at frequency $\nu$ is the frequency itself, i.e.,
$\pfix(\nu) = \nu$; it goes extinct with probability $1-\nu$. For
instance, if a neutral SNV is observed in half of the sequences, it will
fix with a probability of 50\% (see inset in figure
\ref{fig:aftsyn}). The fixation probability of neutral SNVs is
independent of most model assumptions and is only affected if neutral
SNVs are associated preferentially with either high or low fitness
virus.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By analyzing the fate of single nucleotide variants (SNVs) in
longitudinal data of HIV-1 \env{} evolution, we demonstrated selection
against synonymous substitutions in the relatively conserved regions
C2-C4 of the \env{} gene. Computational modeling suggested that these
SNVs have deleterious effects on the order of $0.002$ and that they are
brought to high frequency through linkage to adaptive mutations.
Comparison with biochemical studies of base pairing propensity in an RNA
genome of HIV-1 indicated that these mutations are deleterious, at least
in part, because they disrupt RNA secondary structures. Computer models
of RNA folding predict stable hairpins in these regions that
have been suggested to be functional and termed ``insulating
stems'' \citep{watts_architecture_2009, sanjuan_interplay_2011}.
The weak selection against synonymous variants is compatible with the
negative results of \textit{In vitro} replication assays investigating
the fitness effects of small RNA hairpins in HIV-1
\citep{knoepfel_role_2013}: it would take hundreds of cell culture
passages to detect fitness effects of the order of one per mille. The longitudinal data,
however, spans many years, hence our analysis is able to quantify the
subtle fitness effect of RNA structure within single infections.

>>>>

\begin{abstract}
\noindent
Intrapatient HIV-1 evolution is driven by the adaptive immune system
resulting in rapid change of HIV-1 proteins. When cytotoxic CD8${}^+$ T-cells
or neutralizing antibodies target a new epitope, the virus often escapes via
nonsynonymous mutations that impair recognition. Synonymous mutations do not
affect this interplay and are often assumed to be neutral. We test this
assumption by tracking the abundance of synonymous mutations in 
longitudinal intrapatient data from the \shankaregion{} part of the \env{}
gene. We find that most synonymous variants are lost even though
they often reach high frequencies in the viral population, suggesting a
cost to the virus. Using published data
from SHAPE assays, we find that synonymous mutations that disrupt base pairs
in RNA stems flanking the variable loops of gp120 are more likely to be lost than other
synonymous changes: these RNA hairpins might be important for HIV-1.
Computational modeling indicates that (i) most synonymous mutations in this
genomic region have a selection coefficient of the order of $-0.002$, and
(ii) synonymous variants rise to high frequency by genetic hitchhiking on neighboring beneficial
variants. This weak selection against synonymous substitutions does not
result in a strong pattern of conservation in cross sectional data, but
slows down the rate of evolution considerably. Our findings are consistent with the
notion that large scale patterns of RNA structure are functionally
relevant, while the precise base pairing pattern is not.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
HIV-1 evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via cytotoxic CD8${}^+$ T-cells (CTLs) and neutralizing antibodies
(nAbs)~\citep{rambaut_causes_2004} and facilitated by HIV-1's high mutation rate
~\citep{mansky_lower_1995,abram_nature_2010}. Escape mutations
in epitopes targeted by CTLs are typically observed during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving parts of the HIV-1 genome are the variable
loops V1-V5 in the envelope protein gp120 (V loops), which change to avoid recognition by
nAbs. Escape mutations in \env{}, the gene encoding gp120, spread through the
viral population within a few months.
Consistent with this time scale, it is found that serum from a particular time
typically neutralizes autologous virus extracted more than 3-6 months earlier but not contemporary
virus \citep{richman_rapid_2003}.

@@

Most positions are only transiently variable and variants will either fix or
will be lost -- at least in small samples. Given an SNV is at a certain
frequency $\nu$, the probability of fixation is higher for beneficial
SNVs than for neutral ones; in turn, neutral variants fix more frequently than
deleterious ones. The fixation probability of a
neutral SNV at frequency $\nu$ is the frequency itself, i.e.,
$\pfix(\nu) = \nu$, while it goes extinct with probability $1-\nu$. For
instance, if a neutral SNV is observed in half of the sequences, it will
fix with a probability of 50\% (see inset in figure
\ref{fig:aftsyn}). The fixation probability of neutral SNVs is
independent of most model assumptions and is only affected if neutral
SNVs are associated preferentially with either high or low fitness
virus.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By analyzing the fate of single nucleotide variants (SNVs) in
longitudinal data of HIV-1 \env{} evolution, we demonstrated selection
against synonymous substitutions in the relatively conserved regions
C2-C4. We suggest through computational modeling that these
SNVs have deleterious effects on the order of $0.002$ and that they are
brought to high frequency through linkage to adaptive mutations.
Comparison with biochemical studies of base pairing propensity in an RNA
genome of HIV-1 indicated that these mutations are deleterious, at least
in part, because they disrupt RNA secondary structures. Computer models
of RNA folding predict stable hairpins in these regions that
have been suggested to be functional and termed ``insulating
stems'' \citep{watts_architecture_2009, sanjuan_interplay_2011}.
The weak selection against synonymous variants is compatible with the
negative results of \textit{In vitro} replication assays investigating
the fitness effects of small RNA hairpins in HIV-1
\citep{knoepfel_role_2013}: it would take hundreds of cell culture
passages to detect fitness effects of the order of one per mille. The longitudinal data,
however, spans many years, hence our analysis is able to quantify the
subtle fitness effect of RNA structure within single infections.
